Parameter Recovery Validation

### 2025-06-07 nbs/concepts/parameter-recovery-validation/index.qmd
docs: update parameter recovery validation model description to fix scaling of stimulus onset and duration parameters
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

# PyroVelocity imports
from pyrovelocity.models.modular.factory import (
    create_simple_deterministic_model_with_validation,
    create_legacy_model2,
)
from pyrovelocity.models.modular.components.validation import (
    ParameterRecoveryConfig,
)
from pyrovelocity.io.datasets import bifurcation_14

@@

# Create the simple deterministic model with validation functionality
model = create_simple_deterministic_model_with_validation()

# Configure validation parameters
validation_config = ParameterRecoveryConfig(
    num_parameter_sets=5,       # Number of parameter sets to test
    sample_sizes=[50, 100, 200], # Different sample sizes
    noise_levels=[0.1, 0.2],    # Different noise levels
    inference_algorithm="SVI",   # Use SVI for speed
    num_samples=1000,           # Number of inference samples
    seed=42,                    # Random seed
)

# Run parameter recovery validation using registry-based validation components
report = model.validate_parameter_recovery(validation_config)

print(f"Validation completed!")
print(f"Tested {len(report['parameter_sets'])} parameter sets")
print(f"Overall parameter recovery success rate: {report['summary']['overall_success_rate']:.2%}")

# Quick validation for development
quick_results = model.quick_validation(num_genes=3, num_cells=50, noise_level=0.1)
print(f"Quick validation success rate: {quick_results['summary']['overall_success_rate']:.2%}")

# List available validation components
print(f"Available validation components: {model.list_validation_components()}")
```

@@

The probabilistic graphical model extends Study 2 by adding gene-specific piecewise activation parameters while maintaining the hierarchical latent time structure. For visual clarity, we use $\theta_i$ to represent the vector of piecewise activation parameters $(\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i})$ and $\delta_i$ to represent the activation duration parameter. Note that $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable.

@@

# hyperparameters for gene-specific parameters
pgm.add_node("mu_theta", r"$\mu_{\theta}$", 0.5, 6, fixed=True)
pgm.add_node("sigma_theta", r"$\sigma_{\theta}^2$", 1.5, 6, fixed=True)
pgm.add_node("mu_delta", r"$\mu_{\delta}$", 2.5, 6, fixed=True)
pgm.add_node("sigma_delta", r"$\sigma_{\delta}^2$", 3.5, 6, fixed=True)
pgm.add_node("mu_U0", r"$\mu_{U_0}$", 4.5, 6, fixed=True)
pgm.add_node("sigma_U0", r"$\sigma_{U_0}^2$", 5, 6, fixed=True)

@@

# latent variables for gene-specific parameters
pgm.add_node("theta_i", r"$\theta_i$", 1, 5)
pgm.add_node("delta_i", r"$\delta_i$", 3, 5)
pgm.add_node("U_0i", r"$U_{0i}$", 4.5, 5)

@@

# gene-specific parameter edges
pgm.add_edge("mu_theta", "theta_i", plot_params=edge_params)
pgm.add_edge("sigma_theta", "theta_i", plot_params=edge_params)
pgm.add_edge("mu_delta", "delta_i", plot_params=edge_params)
pgm.add_edge("sigma_delta", "delta_i", plot_params=edge_params)
pgm.add_edge("mu_U0", "U_0i", plot_params=edge_params)
pgm.add_edge("sigma_U0", "U_0i", plot_params=edge_params)
pgm.add_edge("mu_lambda", "lambda_j", plot_params=edge_params)
pgm.add_edge("sigma_lambda", "lambda_j", plot_params=edge_params)

@@

# dynamics edges
pgm.add_edge("theta_i", "u_star_ij", plot_params=edge_params)
pgm.add_edge("theta_i", "s_star_ij", plot_params=edge_params)
pgm.add_edge("delta_i", "u_star_ij", plot_params=edge_params)
pgm.add_edge("delta_i", "s_star_ij", plot_params=edge_params)
pgm.add_edge("t_star_j", "u_star_ij", plot_params=edge_params)
pgm.add_edge("t_star_j", "s_star_ij", plot_params=edge_params)

@@

In the PGM above, we use simplified notation for visual clarity:
- $\theta_i$ represents the vector of piecewise activation parameters $(\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i})$
- $\delta_i$ represents the activation duration parameter $\delta^*_i$ (waiting time from onset to deactivation)
- The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
- The hyperparameters $\mu_{\delta}, \sigma_{\delta}^2$ represent the hyperparameters for $\delta_i$
- $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable

@@

The priors are designed to encode biologically reasonable parameter ranges while eliminating dimensional redundancy:

\begin{align}
    \gamma^*_i &\sim \text{LogNormal}(\log(1.0), 0.5^2) \quad \text{(relative degradation rate)} \\
    R_{\text{on},i} &\sim \text{LogNormal}(\log(2.5), 0.4^2) \quad \text{(activation fold-change)} \\
    t^*_{0,\text{on}i} &\sim \text{Normal}(0.5, 0.8^2) \quad \text{(activation onset time, allows negatives, calibrated)} \\
    \delta^*_i &\sim \text{LogNormal}(\log(0.45), 0.45^2) \quad \text{(activation duration, calibrated for balanced patterns)}
\end{align}

@@

- **$\gamma^*_i \sim \text{LogNormal}(\log(1.0), 0.5^2)$**: Centers relative degradation around 1 (balanced splicing/degradation), with 95% of values between ~0.37 and ~2.7. This covers genes where mRNA is more stable than splicing time ($\gamma^* < 1$) to genes with rapid degradation ($\gamma^* > 1$).

- **$R_{\text{on},i} \sim \text{LogNormal}(\log(2.5), 0.4^2)$**: Centers fold-change at 2.5× during activation, with 95% between ~1.5 and ~4.2. This represents meaningful upregulation while avoiding extreme values that could cause numerical issues.

- **$t^*_{0,\text{on}i} \sim \text{Normal}(0.2, 0.6^2)$**: Centers activation onset at 20% through the process timeline, with 95% between approximately -1.0 and +1.4. **Negative values** represent genes already activated before observation began, enabling modeling of pre-activation scenarios.

- **$\delta^*_i \sim \text{LogNormal}(\log(0.37), 0.35^2)$**: Centers activation duration at 37% of the process timeline, with 95% between ~0.18 and ~0.76. This distribution was optimized to support both transient (< 40%) and sustained (> 50%) activation patterns with sufficient feasibility.

@@

1. **Recover constrained piecewise parameters**: $\gamma^*_i$, $R_{\text{on},i}$, $t^*_{0,\text{on}i}$, $\delta^*_i$
2. **Handle pre-activation scenarios**: Correctly infer negative onset times and their biological implications
3. **Distinguish enhanced expression patterns**: Classify genes across all five patterns including pre-activation
4. **Maintain hierarchical time inference**: Recover population-level temporal parameters
5. **Handle realistic noise levels**: Maintain accuracy with Poisson observation noise
6. **Scale efficiently**: Demonstrate computational advantages of reduced parameter space
7. **Validate fold-change interpretation**: Ensure $R_{\text{on},i}$ estimates match true activation strength
8. **Assess dimensional consistency**: Verify elimination of parameter redundancy issues
9. **Numerical robustness**: Maintain stability across the $\gamma^*_i = 1$ boundary

>>>>

# PyroVelocity imports
from pyrovelocity.models.modular.factory import (
    # create_simple_deterministic_model_with_validation,
    create_legacy_model2,
)
# from pyrovelocity.models.modular.components.validation import (
#     ParameterRecoveryConfig,
# )
from pyrovelocity.io.datasets import bifurcation_14

@@

# Create the simple deterministic model with validation functionality
# model = create_simple_deterministic_model_with_validation()

# # Configure validation parameters
# validation_config = ParameterRecoveryConfig(
#     num_parameter_sets=5,       # Number of parameter sets to test
#     sample_sizes=[50, 100, 200], # Different sample sizes
#     noise_levels=[0.1, 0.2],    # Different noise levels
#     inference_algorithm="SVI",   # Use SVI for speed
#     num_samples=1000,           # Number of inference samples
#     seed=42,                    # Random seed
# )

# # Run parameter recovery validation using registry-based validation components
# report = model.validate_parameter_recovery(validation_config)

# print(f"Validation completed!")
# print(f"Tested {len(report['parameter_sets'])} parameter sets")
# print(f"Overall parameter recovery success rate: {report['summary']['overall_success_rate']:.2%}")

# # Quick validation for development
# quick_results = model.quick_validation(num_genes=3, num_cells=50, noise_level=0.1)
# print(f"Quick validation success rate: {quick_results['summary']['overall_success_rate']:.2%}")

# # List available validation components
# print(f"Available validation components: {model.list_validation_components()}")
```

@@

The probabilistic graphical model extends Study 2 by adding gene-specific piecewise activation parameters while maintaining the hierarchical latent time structure. For visual clarity, we use $\theta_i$ to represent the vector of non-temporal piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$ and $\Delta_i$ to represent the vector of relative temporal gene parameters $(\tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i)$. Note that $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable.

@@

# hyperparameters for gene-specific parameters
pgm.add_node("mu_theta", r"$\mu_{\theta}$", 0.5, 6, fixed=True)
pgm.add_node("sigma_theta", r"$\sigma_{\theta}^2$", 1.5, 6, fixed=True)
pgm.add_node("mu_Delta", r"$\mu_{\Delta}$", 2.5, 6, fixed=True)
pgm.add_node("sigma_Delta", r"$\sigma_{\Delta}^2$", 3.5, 6, fixed=True)
pgm.add_node("mu_U0", r"$\mu_{U_0}$", 4.5, 6, fixed=True)
pgm.add_node("sigma_U0", r"$\sigma_{U_0}^2$", 5, 6, fixed=True)

@@

# latent variables for gene-specific parameters
pgm.add_node("theta_i", r"$\theta_i$", 1, 5)
pgm.add_node("Delta_i", r"$\Delta_i$", 3, 5)
pgm.add_node("U_0i", r"$U_{0i}$", 4.5, 5)

@@

# gene-specific parameter edges
pgm.add_edge("mu_theta", "theta_i", plot_params=edge_params)
pgm.add_edge("sigma_theta", "theta_i", plot_params=edge_params)
pgm.add_edge("mu_Delta", "Delta_i", plot_params=edge_params)
pgm.add_edge("sigma_Delta", "Delta_i", plot_params=edge_params)
pgm.add_edge("mu_U0", "U_0i", plot_params=edge_params)
pgm.add_edge("sigma_U0", "U_0i", plot_params=edge_params)
pgm.add_edge("mu_lambda", "lambda_j", plot_params=edge_params)
pgm.add_edge("sigma_lambda", "lambda_j", plot_params=edge_params)

@@

# time hierarchy edges
pgm.add_edge("T_max_star", "t_star_j", plot_params=edge_params)
pgm.add_edge("t_loc", "t_star_j", plot_params=edge_params)
pgm.add_edge("t_scale", "t_star_j", plot_params=edge_params)
# temporal scaling edge for gene parameters
pgm.add_edge("T_max_star", "Delta_i", plot_params=edge_params)

# dynamics edges
pgm.add_edge("theta_i", "u_star_ij", plot_params=edge_params)
pgm.add_edge("theta_i", "s_star_ij", plot_params=edge_params)
pgm.add_edge("Delta_i", "u_star_ij", plot_params=edge_params)
pgm.add_edge("Delta_i", "s_star_ij", plot_params=edge_params)
pgm.add_edge("t_star_j", "u_star_ij", plot_params=edge_params)
pgm.add_edge("t_star_j", "s_star_ij", plot_params=edge_params)

@@

In the PGM above, we use simplified notation for visual clarity:
- $\theta_i$ represents the vector of **non-temporal** piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$
- $\Delta_i$ represents the vector of **relative temporal** gene parameters $(\tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i)$
- The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
- The hyperparameters $\mu_{\Delta}, \sigma_{\Delta}^2$ represent the collection of hyperparameters for all components of $\Delta_i$
- The absolute timing parameters $t^*_{0,\text{on}i} = T_{M}^* \times \tilde{t}^*_{0,\text{on}i}$ and $\delta^*_i = T_{M}^* \times \tilde{\delta}^*_i$ are computed deterministically
- $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable
- **Key edge**: $T_{M}^* \rightarrow \Delta_i$ represents the scaling relationship for all temporal parameters

@@

The priors are designed to encode biologically reasonable parameter ranges while eliminating dimensional redundancy and cleanly separating temporal from non-temporal parameters:

**Non-temporal gene-specific parameters:**
\begin{align}
    \gamma^*_i &\sim \text{LogNormal}(\log(1.0), 0.5^2) \quad \text{(relative degradation rate)} \\
    R_{\text{on},i} &\sim \text{LogNormal}(\log(2.5), 0.4^2) \quad \text{(activation fold-change)}
\end{align}

**Relative temporal gene-specific parameters:**
\begin{align}
    \tilde{t}^*_{0,\text{on}i} &\sim \text{Normal}(0.5, 0.8^2) \quad \text{(relative activation onset time, allows negatives)} \\
    \tilde{\delta}^*_i &\sim \text{LogNormal}(\log(0.45), 0.45^2) \quad \text{(relative activation duration)}
\end{align}

**Absolute temporal parameters (deterministic transformations):**
\begin{align}
    t^*_{0,\text{on}i} &= T_{M}^* \times \tilde{t}^*_{0,\text{on}i} \quad \text{(absolute activation onset time)} \\
    \delta^*_i &= T_{M}^* \times \tilde{\delta}^*_i \quad \text{(absolute activation duration)}
\end{align}

@@

**Non-temporal parameters:**
- **$\gamma^*_i \sim \text{LogNormal}(\log(1.0), 0.5^2)$**: Centers relative degradation around 1 (balanced splicing/degradation), with 95% of values between ~0.37 and ~2.7.

- **$R_{\text{on},i} \sim \text{LogNormal}(\log(2.5), 0.4^2)$**: Centers fold-change at 2.5× during activation, with 95% between ~1.5 and ~4.2.

**Relative temporal parameters:**
- **$\tilde{t}^*_{0,\text{on}i} \sim \text{Normal}(0.5, 0.8^2)$**: Centers relative activation onset at 50% through the process timeline, with 95% between approximately -1.1 and +2.1. **Negative values** enable pre-activation scenarios.

- **$\tilde{\delta}^*_i \sim \text{LogNormal}(\log(0.45), 0.45^2)$**: Centers relative activation duration at 45% of the process timeline, with 95% between ~0.18 and ~1.13.

**Temporal scaling transformation:**
- **Absolute onset**: $t^*_{0,\text{on}i} = T_{M}^* \times \tilde{t}^*_{0,\text{on}i}$ scales relative onset to absolute time
- **Absolute duration**: $\delta^*_i = T_{M}^* \times \tilde{\delta}^*_i$ scales relative duration to absolute time
- **Dimensional consistency**: All temporal comparisons now use consistently scaled absolute times

**Parameter interpretation benefits:**
- **Separation of concerns**: Non-temporal kinetics ($\theta_i$) vs temporal patterns ($\Delta_i$)
- **Scale invariance**: Relative temporal patterns are independent of absolute time scale
- **Biological realism**: Global time scale $T_{M}^*$ sets the pace for the entire process

@@

1. **Recover constrained piecewise parameters**: $\gamma^*_i$, $R_{\text{on},i}$, $t^*_{0,\text{on}i}$, $\delta^*_i$
2. **Handle pre-activation scenarios**: Correctly infer negative onset times and their biological implications
3. **Distinguish enhanced expression patterns**: Classify genes across all five patterns including pre-activation
4. **Maintain hierarchical time inference**: Recover population-level temporal parameters
5. **Handle realistic noise levels**: Maintain accuracy with Poisson observation noise
6. **Scale efficiently**: Demonstrate computational advantages of reduced parameter space
7. **Validate fold-change interpretation**: Ensure $R_{\text{on},i}$ estimates match true activation strength
8. **Assess dimensional consistency**: Verify elimination of parameter redundancy issues
9. **Validate temporal parameter scaling**: Ensure $T_{M}^*$ scaling correctly propagates to absolute timing parameters
10. **Assess parameter separation**: Verify clean separation between kinetic and temporal parameter inference
